---
title: "Your Project Title"
author: 
  - "Alex Delamer"
  - "Duc Hoang"
  - "Aatish Prasad"
  - "Kishori Akola"
date: today
format: 
  html:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-section: true
bibliography: references.bib
csl: apa.csl
lightbox: true
execute:
  echo: false
  output: true
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lubridate)
library(janitor)
library(knitr)
library(scales)
library(kableExtra)
```

## Introduction

### Research Design and Purpose

> **Purpose**: to examine the sales performance of a UK-based online retail business over a 13-month period from December 2018 to December 2019.

Specifically, the analysis investigates transaction volume and revenue trends over time, and evaluates whether performance differs between the United Kingdom and international markets.

This study first adopts a descriptive analytical approach, utilising transactional records to summarise and visualise patterns in the data. Furthermore, further business insights are further investigated to identify key values that would inform future operational and marketing decisions.

```{r data-load}
#| echo: false
#| message: false
#| results: false
## Data Loading
sales_data <- read_csv("data/Sales Transaction v.4a.csv")

glimpse(sales_data)

```

## Methodology

### Data Collection and Sources

This report utilizes secondary data sourced from *Kaggle*, a prominent open-source data repository (@ramos_e-commerce_nodate). The dataset comprises one year of historical sales transactions from Dec 2018 to Dec 2019 from a UK-based online retailer. This retailer is headquartered in London, specializing in gifts and homewares for both adults and children.

The business operates on a global scale and serves a dual customer base. Operations include **both direct-to-consumer (B2C) sales**, where individuals purchase for personal use, **and business-to-business (B2B) transactions**, where small businesses purchase items in bulk for retail distribution. Consequently, the transaction data is expected to exhibit diverse purchasing patterns, reflecting both individual consumer behavior and wholesale buying cycles.

Structurally, the dataset contains **536,350 observations** and **8 variables**. Each record represents the quantity of a single product sold as part of a transaction; a transaction can have multiple products.

### Preparing Dataset for Analysis

To ensure data integrity and facilitate robust analysis, the raw dataset underwent a systematic cleaning and transformation process. Initially, variable names were standardized for consistency, and exact duplicate records were identified and removed to prevent observation bias and the double-counting of revenue.

-   **Standardization & Deduplication**: Column headers were standardized, and duplicate records were removed from the dataset.

-   **Financial Feature Engineering**: A discrete `revenue` variable was calculated for each line item (Price × Quantity).

-   **Temporal Decomposition**: The original date string was parsed into a datetime format and decomposed into `year`, `month`, and `year_month` fields to facilitate seasonal and time-series analysis.

-   **Transaction Classification**: A `transaction_type` indicator was created to isolate legitimate sales from returns. Any record with a negative quantity or a transaction ID starting with **"C"** was flagged as a `Cancelled/Return`.

Following the deduplication and feature engineering phases, the final prepared dataset consists of 531,150 observations across 13 variables.

### Analysing Sales Trends over Time

In order to identify trends in both number of transactions and total revenue for the UK and non-UK markets, all transactions in the dataset were first assigned one of these two regions. This allowed for the creation of @fig-overview, a collection of time-series plots for transactions and revenue divided by region.

These plots allow for the easy visual identification of market trends across the 13-month period we have data for.

```{r data-cleaning}

sales_clean <- sales_data |>
  clean_names() |>
  mutate(
    date = mdy(date),
    revenue = price * quantity,
    year = year(date),
    month = month(date, label = TRUE),
    year_month = floor_date(date, "month"),

    # flagging order cancellations
    transaction_type = if_else(
      quantity < 0 | str_starts(transaction_no, "C"),
      "Cancelled/Return",
      "Sale"
    )
  ) |> 
  distinct() # deduplication
```

```{r}
#| fig-cap: "Monthly Transaction and Revenue in UK vs non-UK countries in 2018-2019 period"
#| label: fig-overview


sales_with_region <- sales_clean |>
  filter(transaction_type == "Sale") |>
  mutate(
    region     = ifelse(country == "United Kingdom", "UK", "Non-UK"),
    year_month = as.Date(year_month)
  )

monthly_long <- sales_with_region |>
  group_by(year_month, region) |>
  summarise(
    `No. of Transactions` = n_distinct(transaction_no),
    `Total Revenue (£M)`  = sum(revenue, na.rm = TRUE) / 1e6,
    .groups = "drop"
  ) |>
  pivot_longer(
    cols      = c(`No. of Transactions`, `Total Revenue (£M)`),
    names_to  = "metric",
    values_to = "value"
  )

ggplot(monthly_long, aes(x = year_month, y = value, fill = region)) +
  geom_col(alpha = 0.8) +
  facet_grid(metric ~ region, scales = "free_y") +
  scale_x_date(date_labels = "%b\n%y", date_breaks = "1 month") +
  scale_y_continuous(labels = label_comma(accuracy = 0.1)) +
  scale_fill_manual(values = c("UK" = "#4C72B0", "Non-UK" = "#DD8453")) +
  labs(
    title    = "Figure 1. Monthly Transactions and Revenue: UK vs Non-UK",
    subtitle = "Sales only  ·  Dec 2018 – Dec 2019",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "none",
    strip.text       = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "grey92", colour = NA),
    axis.text.x      = element_text(size = 7),
    panel.spacing    = unit(1, "lines"),
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(colour = "grey40", size = 10)
  )
```

## Results

### Number of orders and total Volume by month

As presented in @fig-overview, The UK market consistently dominated in both number of transactions and revenue throughout the recorded period. A pronounced upward trend is observed from September 2019 onwards, with transactions and revenue in the UK peaking in November 2019 at nearly 2500 orders and approximately £6.8M respectively, likely driven by seasonal demand ahead of the holiday period.

The Non-UK market, while significantly smaller in absolute scale, demonstrated a gradual growth trajectory across 2019 with the upward trend merely mirror UK market. This suggests a slow but steady expansion of the international customer base.

### General Quantitative metrics

As seen in @fig-overview, both the UK and Non-UK markets reached their peak transaction volumes and revenue towards the end of 2019. Referencing @tbl-overview, the UK reached 2487 transactions and £6.81M in revenue in November 2019, while Non-UK countries peaked at 266 transactions in the same month. This Q4 surge is consistent with seasonal holidays like Christmas and New Year.

Notably, Non-UK maximum revenue occurred in October 2019 (£1.37M) rather than November despite November having more transactions, suggesting a higher average basket value per order in October.

The minimum values for both regions fall in December 2019 across all metrics. This might not be a genuine business downturn but rather a data artefact, as December 2019 represents an incomplete month in the dataset.

```{r}
#| tbl-cap: "Monthly Transaction and Revenue in UK vs non-UK countries in 2018-2019 period"
#| label: tbl-overview

overview_table <- sales_with_region |> 
  group_by(year_month, region) |>
  summarise(
    Transactions = n_distinct(transaction_no),
    Revenue = sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(region) |>
  summarise(
    `Total Transactions`      = sum(Transactions),
    `Mean Transactions/Month` = round(mean(Transactions), 0),
    `Min Transactions/Month`  = min(Transactions),
    `Max Transactions/Month`  = max(Transactions),
    `Total Revenue (£)`       = round(sum(Revenue), 2),
    `Mean Revenue/Month (£)`  = round(mean(Revenue), 2),
    `Min Revenue/Month (£)`   = round(min(Revenue), 2),
    `Max Revenue/Month (£)`   = round(max(Revenue), 2),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(
    cols = -region,
    names_to = "Metric",
    values_to = "Value"
  ) |>
  tidyr::pivot_wider(
    names_from = region,
    values_from = Value
  ) |> 
  mutate(
    across(
      -Metric,
      ~ ifelse(
        grepl("Revenue", Metric),
        format(round(.x, 2), big.mark = ",", nsmall = 2),
        format(round(.x, 0), big.mark = ",", nsmall = 0)
      )
    )
  )

overview_table |>
  kable(
    align   = c("l", rep("r", 8))
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width        = FALSE,
    position          = "center",
    font_size         = 13
  ) |>
  add_header_above(
    c(" " = 1,
      "Region" = 2)
  ) |>
  row_spec(0, bold = TRUE, background = "#4C72B0", color = "white") |>
  column_spec(1, bold = TRUE)
```

```{r}
#| label: fig-non-uk-figures
#| fig-cap: "Transactions and Revenue for Non-UK Countries"

transaction_summary <- sales_clean |>
  group_by(transaction_no) |>
  summarise(
    transaction_revenue = sum(revenue),
    country = first(country),
    year_month = first(year_month),
    .groups = "drop"
  )

country_summary <- sales_clean |>
  group_by(country) |>
  summarise(
    Transactions = n_distinct(transaction_no),
    Revenue = sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    `Avg Revenue per Transaction (£)` = round(Revenue / Transactions, 2)
  ) |>
  arrange(desc(Revenue))

country_summary <- sales_clean |>
  filter(country != "United Kingdom") |>
  group_by(country) |>
  summarise(
    Transactions = n_distinct(transaction_no),
    Revenue = sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = c(Transactions, Revenue),
    names_to = "Metric",
    values_to = "Value"
  )

country_summary |>
  ggplot(aes(x = reorder(country, Value), y = Value)) +
  geom_col() +
  facet_wrap(~ Metric, scales = "free_x") +
  coord_flip() +
  labs(
    x = "Country",
    y = NULL
  ) +
  theme_minimal()
```


## References

---
nocite: |
  @tidyverse-cite, @lubridate-cite, @knitr-cite, @janitor-cite, @scales-cite, @kableExtra-cite 
---

::: {#refs}

:::



