---
title: "UK E-commerce Sales Performance Analysis 2018–2019"
author: 
  - "Alex Delamer"
  - "Duc Hoang"
  - "Aatish Prasad"
  - "Kishori Akola"
date: today
format: 
  html:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-section: true
bibliography: references.bib
csl: apa.csl
lightbox: true
execute:
  echo: false
  output: true
  warning: false
  message: false
---

::: callout-important
Please run `renv::restore()` in the R console before attempting to render this document.
:::

```{r setup}
#| output: false

library(tidyverse)
library(lubridate)
library(janitor)
library(knitr)
library(scales)
library(kableExtra)
```

## Executive Summary

This report analyses the retail transaction data of a UK based e-retailer selling gifts and house ware for all ages, to identify sales and revenue patterns across the United Kingdom and international markets during the Jan 2019 - Dec 2019 period. The data set was cleaned by removing duplicate records and creating new variables to support analysis. The primary focus of this report was to compare transaction volume and the revenue generated within the UK and in the international market, with a secondary focus on identifying countries outside the UK that may represent viable growth opportunities. Results indicated that while the majority of transaction volume and revenue was generated by the UK, a few countries, such as the Netherlands and Ireland (EIRE), had a disproportionately higher revenue against transaction volume, which may merit further investigation.

## Introduction

### Research Design and Purpose

> **Purpose**: examine the sales performance of a UK-based online retail business over a 13-month period from December 2018 to December 2019.

Specifically, the analysis investigates transaction volume and revenue trends over time, and evaluates whether performance differs between the United Kingdom and international markets.

This study first adopts a descriptive analytical approach, utilising transactional records to summarise and visualise patterns in the data. Furthermore, further business insights are further investigated to identify key values that would inform future operational and marketing decisions.

```{r data-load}
sales_data <- read_csv("data/Sales Transaction v.4a.csv")
```

## Methodology

### Data Collection and Sources

This report utilizes secondary data sourced from *Kaggle*, a prominent open-source data repository (@ramos_e-commerce_nodate). The dataset comprises one year of historical sales transactions from Dec 2018 to Dec 2019 from a UK-based online retailer. This retailer is headquartered in London, specializing in gifts and homewares for both adults and children.

The business operates on a global scale and serves a dual customer base. Operations include **both direct-to-consumer (B2C) sales**, where individuals purchase for personal use, **and business-to-business (B2B) transactions**, where small businesses purchase items in bulk for retail distribution. Consequently, the transaction data is expected to exhibit diverse purchasing patterns, reflecting both individual consumer behavior and wholesale buying cycles.

Structurally, the dataset contains **536,350 observations** and **8 variables**. Each record represents the quantity of a single product sold as part of a transaction; a transaction can have multiple products.

```{r}
#| tbl-cap: "Raw Dataset Variables Descriptions"
#| label: tbl-raw-data


tibble(
  Variable = c("TransactionNo", "Date", "ProductNo", "ProductName", "Price", "Quantity", "CustomerNo", "Country"),
  Type = c("Character", "Character", "Character", "Character", "Numeric", "Numeric", "Numeric", "Character"),
  Description = c("Unique transaction identifier (prefix 'C' = cancellation)",
                       "Date of transaction",
                       "Unique product identifier",
                       "Name of product sold",
                       "Unit price in GBP (£)",
                       "Units sold (negative values indicate returns)",
                       "Unique customer identifier",
                       "Country of the customer")
) %>%
  kable(
    align   = c("l", "l", "l")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    font_size         = 13
  ) %>%
  row_spec(0, bold = TRUE, background = "#4C72B0", color = "white") %>%
  column_spec(1, bold = TRUE, monospace = TRUE)
```

### Preparing Dataset for Analysis

To ensure data integrity and facilitate robust analysis, the raw dataset underwent a systematic cleaning and transformation process. Initially, variable names were standardized for consistency, and exact duplicate records were identified and removed to prevent observation bias and the double-counting of revenue.

-   **Standardization & Deduplication**: Column headers were standardized, and duplicate records were removed from the dataset.

-   **Financial Feature Engineering**: A discrete `revenue` variable was calculated for each line item (Price × Quantity).

-   **Temporal Decomposition**: The original date string was parsed into a datetime format and decomposed into `year`, `month`, and `year_month` fields to facilitate seasonal and time-series analysis.

-   **Transaction Classification**: A `transaction_type` indicator was created to isolate legitimate sales from returns. Any record with a negative quantity or a transaction ID starting with **"C"** was flagged as a `Cancelled/Return`.

Following the deduplication and feature engineering phases, the final prepared dataset consists of 531,150 observations across 13 variables.

### Analysing Sales Trends over Time

In order to identify trends in both number of transactions and total revenue for the UK and non-UK markets, all transactions in the dataset were first assigned one of these two regions. This allowed for the creation of @fig-overview, a collection of time-series plots for transactions and revenue divided by region.

These plots allow for the easy visual identification of market trends across the 13-month period we have data for.

```{r data-cleaning}

sales_clean <- sales_data |>
  clean_names() |>
  mutate(
    date = mdy(date),
    revenue = price * quantity,
    year = year(date),
    month = month(date, label = TRUE),
    year_month = floor_date(date, "month"),
    # flagging order cancellations
    transaction_type = if_else(
      quantity < 0 | str_starts(transaction_no, "C"),
      "Cancelled/Return",
      "Sale"
    )
  ) |> 
  distinct() # deduplication

```

```{r}
#| fig-cap: "Monthly Transaction and Revenue in UK vs non-UK countries in 2018-2019 period"
#| label: fig-overview

sales_with_region <- sales_clean |>
  filter(transaction_type == "Sale") |>
  mutate(
    region     = ifelse(country == "United Kingdom", "UK", "Non-UK"),
    year_month = as.Date(year_month)
  )

monthly_long <- sales_with_region |>
  group_by(year_month, region) |>
  summarise(
    `No. of Transactions` = n_distinct(transaction_no),
    `Total Revenue (£M)`  = sum(revenue, na.rm = TRUE) / 1e6,
    .groups = "drop"
  ) |>
  pivot_longer(
    cols      = c(`No. of Transactions`, `Total Revenue (£M)`),
    names_to  = "metric",
    values_to = "value"
  )

monthly_uk_nonuk <- ggplot(monthly_long, aes(x = year_month, y = value, fill = region)) +
  geom_col(alpha = 0.8) +
  facet_grid(metric ~ region, scales = "free_y") +
  scale_x_date(date_labels = "%b\n%y", date_breaks = "1 month") +
  scale_y_continuous(labels = label_comma(accuracy = 0.1)) +
  scale_fill_manual(values = c("UK" = "#4C72B0", "Non-UK" = "#DD8453")) +
  labs(
    title    = "Monthly Transactions and Revenue: UK vs Non-UK",
    subtitle = "Sales only  ·  Dec 2018 – Dec 2019",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "none",
    strip.text       = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "grey92", colour = NA),
    axis.text.x      = element_text(size = 7),
    panel.spacing    = unit(1, "lines"),
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(colour = "grey40", size = 10)
  )

ggsave(monthly_uk_nonuk, filename = "media/monthly_uk_nonuk.png")

monthly_uk_nonuk
```

## Results

### Number of orders and total Volume by month

As presented in @fig-overview, The UK market consistently dominated in both number of transactions and revenue throughout the recorded period. A pronounced upward trend is observed from September 2019 onwards, with transactions and revenue in the UK peaking in November 2019 at nearly 2500 orders and approximately £6.8M respectively, likely driven by seasonal demand ahead of the holiday period.

The Non-UK market, while significantly smaller in absolute scale, demonstrated a gradual growth trajectory across 2019 with the upward trend merely mirror UK market. This suggests a slow but steady expansion of the international customer base.

@fig-non-uk-figures shows the top non-UK countries by revenue amount, and also hints at an interesting idea; that the retailer is performing very well in some countries despite relatively low numbers of transactions. This will be covered in more depth in @sec-basket-values.

```{r}
#| label: fig-non-uk-figures
#| fig-cap: "Transactions and Revenue for Non-UK Countries"

transaction_summary <- sales_clean |>
  group_by(transaction_no) |>
  summarise(
    transaction_revenue = sum(revenue),
    country = first(country),
    year_month = first(year_month),
    .groups = "drop"
  )

country_summary <- sales_clean |>
  group_by(country) |>
  summarise(
    Transactions = n_distinct(transaction_no),
    Revenue = sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    `Avg Revenue per Transaction (£)` = round(Revenue / Transactions, 2)
  ) |>
  arrange(desc(Revenue))

country_summary <- sales_clean |>
  filter(country != "United Kingdom") |>
  group_by(country) |>
  summarise(
    Transactions = n_distinct(transaction_no),
    Revenue = sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = c(Transactions, Revenue),
    names_to = "Metric",
    values_to = "Value"
  )

country_summary_plot <- country_summary |>
  ggplot(aes(x = reorder(country, Value), y = Value)) +
  geom_col() +
  facet_wrap(~ Metric, scales = "free_x") +
  coord_flip() +
  labs(
    title = "Revenue against Number of Transactions by Country",
    x = "Country",
    y = NULL
  ) +
  theme_minimal()

ggsave(country_summary_plot, filename = "media/country_summary_plot.png")

country_summary_plot
```

### General Quantitative metrics

As seen in @fig-overview, both the UK and Non-UK markets reached their peak transaction volumes and revenue towards the end of 2019. Referencing @tbl-overview, the UK reached 2487 transactions and £6.81M in revenue in November 2019, while Non-UK countries peaked at 266 transactions in the same month. This Q4 surge is consistent with seasonal holidays like Christmas and New Year.

Notably, Non-UK maximum revenue occurred in October 2019 (£1.37M) rather than November despite November having more transactions, suggesting a higher average basket value per order in October.

The minimum values for both regions fall in December 2019 across all metrics. This might not be a genuine business downturn but rather a data artefact, as December 2019 represents an incomplete month in the dataset.

```{r}
#| tbl-cap: "Monthly Transaction and Revenue in UK vs non-UK countries in 2018-2019 period"
#| label: tbl-overview

overview_table <- sales_with_region |> 
  group_by(year_month, region) |>
  summarise(
    Transactions = n_distinct(transaction_no),
    Revenue = sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(region) |>
  summarise(
    `Total Transactions`      = sum(Transactions),
    `Mean Transactions/Month` = round(mean(Transactions), 0),
    `Min Transactions/Month`  = min(Transactions),
    `Max Transactions/Month`  = max(Transactions),
    `Total Revenue (£)`       = round(sum(Revenue), 2),
    `Mean Revenue/Month (£)`  = round(mean(Revenue), 2),
    `Min Revenue/Month (£)`   = round(min(Revenue), 2),
    `Max Revenue/Month (£)`   = round(max(Revenue), 2),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(
    cols = -region,
    names_to = "Metric",
    values_to = "Value"
  ) |>
  tidyr::pivot_wider(
    names_from = region,
    values_from = Value
  ) |> 
  mutate(
    across(
      -Metric,
      ~ ifelse(
        grepl("Revenue", Metric),
        format(round(.x, 2), big.mark = ",", nsmall = 2),
        format(round(.x, 0), big.mark = ",", nsmall = 0)
      )
    )
  )

overview_table |>
  kable(
    align   = c("l", rep("r", 8))
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width        = FALSE,
    position          = "center",
    font_size         = 13
  ) |>
  add_header_above(
    c(" " = 1,
      "Region" = 2)
  ) |>
  row_spec(0, bold = TRUE, background = "#4C72B0", color = "white") |>
  column_spec(1, bold = TRUE)
```

### Countries' basket value {#sec-basket-values}

@fig-basket-value presents basket size statistics across top 10 countries with the biggest basket and the UK. Netherlands, Australia and Singapore recorded the highest mean basket values at £22888, £16055 and £15870 respectively. Notably, these countries achieved these values from a very small number of transactions, especially Singapore had only 4 transactions, this pattern could indicate wholesale or B2B purchasing behaviour rather than individual retail orders.

```{r}
#| fig-cap: "Metrics for Basket Size per transaction by country in 2018-2019 period"
#| label: fig-basket-value

basket_country <- sales_clean |>
  filter(transaction_type == "Sale") |> 
  group_by(country, transaction_no) |>
  summarise(basket_value = sum(revenue, na.rm = TRUE), .groups = "drop") |>
  group_by(country) |>
  summarise(
    `Transactions`           = n(),
    `Mean Basket (£)`        = round(mean(basket_value), 2),
    .groups = "drop"
  ) |>
  arrange(desc(`Mean Basket (£)`)) |>
  rename(Country = country)

top10 <- basket_country |> slice_max(`Mean Basket (£)`, n = 10)
uk_row <- basket_country |> filter(Country == "United Kingdom")

basket_size_plot <- bind_rows(top10, uk_row) |>
  distinct(Country, .keep_all = TRUE) |>
  mutate(
    Country   = fct_reorder(Country, `Mean Basket (£)`),
    highlight = case_when(
      Country %in% c("Netherlands", "Australia", "Singapore") ~ "High-Value",
      Country == "United Kingdom"                ~ "UK",
      TRUE                                       ~ "Other"
    )
  ) |>
  pivot_longer(                                        # ← key change
    cols     = c(`Mean Basket (£)`, `Transactions`),
    names_to  = "metric",
    values_to = "value"
  ) |>
  ggplot(aes(x = value, y = Country, fill = highlight)) +
  geom_col(alpha = 0.85) +
  geom_text(
    aes(label = ifelse(metric == "Mean Basket (£)",
                       paste0("£", comma(round(value))),
                       comma(value))),
    hjust = -0.1, size = 3.2
  ) +
  scale_fill_manual(
    values = c("High-Value" = "#e8a838",
               "UK"         = "#4C72B0",
               "Other"      = "#a8c4d4")
  ) +
  scale_x_continuous(
    labels = label_comma(),
    expand = expansion(mult = c(0, 0.25))
  ) +
  facet_wrap(~ metric, scales = "free_x") +           # ← free_x = each panel own scale
  labs(
    title    = "Mean Basket Size and Number of Transactions by Country",
    subtitle = "Top 10 countries by basket size + UK as reference · Sales only · Dec 2018 – Dec 2019",
    x        = NULL,
    y        = NULL,
    fill     = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position    = "top",
    strip.text         = element_text(face = "bold", size = 11),
    strip.background   = element_rect(fill = "grey92", colour = NA),
    plot.title         = element_text(face = "bold", size = 13),
    panel.grid.major.y = element_blank()
  )

ggsave(basket_size_plot, filename = "media/basket_size_plot.png")

basket_size_plot
```

### Cancelled orders insight

@fig-cancellation shows that the USA has the highest cancellation rate at 29.63%, followed by Malta (9.4%) and Japan (9.16%), all well above the UK benchmark of 1.52%. However, despite its low rate, the UK incurs the largest revenue loss at £2509.3K due to its transaction number. Among Non-UK markets, EIRE (Ireland) records the highest revenue lost at £52.8K, making it the most financially impactful cancellation market outside the UK.

```{r}
#| fig-cap: "Cancellation rate and Revenue lost by each Country"
#| label: fig-cancellation


cancel_summary <- sales_clean |>
  group_by(country) |>
  summarise(
    total        = n(),
    cancelled    = sum(transaction_type == "Cancelled/Return"),
    `Cancel Rate (%)` = round(cancelled / total * 100, 2),
    `Revenue Lost (£K)` = round(abs(sum(
      revenue[transaction_type == "Cancelled/Return"],
      na.rm = TRUE)) / 1000, 1),
    .groups = "drop"
  ) |>
  filter(total > 100) |>                             
  arrange(desc(`Cancel Rate (%)`)) |>
  rename(Country = country) |>
  select(Country, `Cancel Rate (%)`, `Revenue Lost (£K)`)

cancel_summary_plot <- cancel_summary |>
  filter(`Cancel Rate (%)` > 0) |> 
  mutate(
    Country   = fct_reorder(Country, `Cancel Rate (%)`),
    highlight = case_when(
      Country == "United Kingdom"  ~ "UK",
      `Cancel Rate (%)` > 3        ~ "High Risk",
      TRUE                         ~ "Other"
    )
  ) |>
  pivot_longer(
    cols      = c(`Cancel Rate (%)`, `Revenue Lost (£K)`),
    names_to  = "metric",
    values_to = "value"
  ) |>
  ggplot(aes(x = value, y = Country, fill = highlight)) +
  geom_col(alpha = 0.85) +
  geom_text(
    aes(label = ifelse(metric == "Cancel Rate (%)",
                       paste0(value, "%"),
                       paste0("£", value, "K"))),
    hjust = -0.1, size = 3.2
  ) +
  scale_fill_manual(
    values = c("High Risk" = "#C44E52",
               "UK"        = "#4C72B0",
               "Other"     = "#a8c4d4"),
    labels = c("High Risk" = "High Risk (>3%)",
               "UK"        = "United Kingdom",
               "Other"     = "Other Countries")
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.3))) +
  facet_wrap(~ metric, scales = "free_x") +
  labs(
    title   = "Cancellation Rate and Revenue Lost by Country",
    subtitle = "Countries with fewer than 100 records excluded  ·  Dec 2018 – Dec 2019",
    caption = "Source: Retail transactions dataset",
    x       = NULL,
    y       = NULL,
    fill    = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position    = "top",
    strip.text         = element_text(face = "bold", size = 11),
    strip.background   = element_rect(fill = "grey92", colour = NA),
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(colour = "grey40", size = 9),
    panel.grid.major.y = element_blank()
  )

ggsave(cancel_summary_plot, filename = "media/cancel_summary_plot.png")

cancel_summary_plot
```

## Discussion

The results indicates that this business is operationally strong in its domestic market but has significant untapped potential internationally.

### General market is Seasonal

The dominant Q4 seasonality observed in @fig-overview suggests the business is **heavily reliant on the festive season period** to drive annual performance. While this concentration generates strong peak revenue, it also **exposes the business to seasonal risk**.

### High value international B2B market

The basket size analysis in @fig-basket-value reveals a more complex international picture. Despite lower transaction volumes, most **Non-UK customers spend nearly exponentially more per order comparing to the UK**. Countries such as the Netherlands, Singapore, and Australia demonstrate average basket sizes of approximately £16000, mostly indicate that they belong to the **B2B customer segment**. Rather than casual retail consumers, these clients are bulk purchasers placing infrequent, high-value orders, making them disproportionately valuable relative to their transaction volume.

### Barrier regarding international sales

However, @fig-cancellation highlights a key operational challenge in international markets. The cancellation rate at the USA (29.6%), Japan (9.2%) and Germany (4.1%) all exceed the UK benchmark of 1.52%, suggesting that international fulfillment reliability remains inconsistent. While EIRE represents the largest Non-UK revenue loss at £52.8K, the high-rate markets like the USA likely reflect **demand uncertainty** or **logistics hindrance** rather than product dissatisfaction.

## Recommendation

**Reduce seasonal dependency:** The business should invest in mid-year promotional campaigns, particularly in months of February and April while they showed the modest number, to reduce the financial risk of a single underperforming holiday season.

**Formalise and grow the international wholesale channel:** The high basket sizes recorded in the Netherlands, Australia and Singapore signal an existing wholesale customer base. Rather than treating these as ordinary transactions, the business should *develop a dedicated B2B programme with volume-based pricing, account management and tailored outreach*.

**Address international fulfillment gaps before scaling:** The elevated cancellation rates in some countries suggest that the operational infrastructure for international orders is not yet mature. Before investing further in these markets, the business should investigate the *root causes of cancellations: whether driven by shipping delays, product mismatches, or payment friction*. The Netherlands, with a near-zero cancellation rate of 0.17%, offers a useful operational benchmark for what reliable international fulfillment looks like.

## Conclusion

Ultimately, we find that the vast majority of transaction activity originates in the United Kingdom, and therefore the retailer's primary focus should remain here. At the same time, there should be a concerted effort to develop the international markets and supply chain. In particular, the retailer should focus on expanding its B2B operations, as our analysis suggests that a significant portion of international revenue seems to stem from business customers ordering bulk orders.

## References

---
nocite: |
  @tidyverse-cite, @lubridate-cite, @knitr-cite, @janitor-cite, @scales-cite, @kableExtra-cite, @renv-cite
---

::: {#refs}
:::
